---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Sex health clinics in Toronto

<!-- badges: start -->
<!-- badges: end -->

This repository holds information about sex health clinics in Toronto. It supports a student project for ENVSOCTY 4GA3 (Applied Spatial Statistics) by:

- Victoria Baginski  
- Eva Boomsma  
- Peri Juskiw
- Samantha Kirtz  
- Chantelle Lobo  
- Helena Muirhead-Hunt  
- Eva Novoselac  
- Audreana Rossi  

The students have collected information about clinics in Toronto, and have the Dissemination Areas (DAs). They shared the DA centroids and the location of the clinics.

I obtained the road network in Toronto from [BBBike](https://download.bbbike.org/osm/bbbike/Toronto), and use {[r5r](https://ipeagit.github.io/r5r/index.html)} to calculate driving times from DA centroids to each clinic. 

The file with the road network is not shared on GitHub (it is a large file), so if you wish to replicate the routing calculations you need to obtain a copy and place in folder `data-raw/r5_graph/`. The file must be in `osm.pbf` format, which is what {r5r} uses.

The notebook with the routing is in folder `data-raw\01-OSM-Network-and-Routing`. You can check it for details.

The following data objects are available:

- `data/clinics.rda`: a simple features table with the location of the clinics.  
- `data/da_centroids.rda`: a simple features table with the centroids of the Dissemination Areas (DAs) in Toronto.
- `data/ttm_driva_da.rda`: a simple features table with Toronto's Dissemination Areas (DAs).  
- `data/ttm_driva_da.rda`: a data frame with driving times from DA centroid to clinic. The travel time is in minutes.

```{r load-packages, include = FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(sf) # Simple Features for R
```

```{r load-data, include=FALSE}
load("data/ttm_drive_da.rda")
load("data/toronto_das.rda")
```

```{r calculate-impedance, include=FALSE}
# Calculate impedance; use the formula used by Paez, Higgins, and Vivona (2019). Check equation in page 27.
ttm_drive_da <- ttm_drive_da |>
  mutate(f_tt = case_when(travel_time <= 5 ~ 0.946,
                           travel_time > 5 & travel_time <= 10 ~ 0.801,
                           travel_time > 10 & travel_time <= 15 ~ 0.607,
                           travel_time > 15 & travel_time <= 20 ~ 0.411,
                           travel_time > 20 & travel_time <= 30 ~ 0.135,
                           travel_time > 30 & travel_time <= 45 ~ 0.011,
                           travel_time > 45 ~ 0.000))
```

```{r calculate-accessibility, include=FALSE}
acc_sex <- ttm_drive_da |>
  group_by(DAUID) |>
  summarize(acc_sex = sum(f_tt),
            .groups = "drop")
```

```{r summary-accessibility, include=FALSE}
summary(acc_sex)
```

```{r join-accessibility-to-DAs, include=FALSE}
toronto_das <- toronto_das |>
  left_join(acc_sex,
            by = "DAUID")
```

```{r plot-accessibility, include=FALSE}
ggplot() +
  geom_sf(data = toronto_das,
          aes(fill = acc_sex),
          color = NA) +
  scale_fill_distiller(name = "Accessibility",
                       palette = "Reds", 
                       direction = 1,
                       na.value = "white") +
  labs(subtitle = "Accessibility to sex health clinics by DA") +
  theme_minimal()
```



