---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Sex health clinics in Toronto

<!-- badges: start -->
<!-- badges: end -->

This repository holds information about sex health clinics in Toronto. It supports a student project for ENVSOCTY 4GA3 (Applied Spatial Statistics) by:

- Victoria Baginski  
- Eva Boomsma  
- Peri Juskiw
- Samantha Kirtz  
- Chantelle Lobo  
- Helena Muirhead-Hunt  
- Eva Novoselac  
- Audreana Rossi  

The students have collected information about clinics in Toronto, and have the Dissemination Areas (DAs). They shared the DA centroids and the location of the clinics. Additional information includes population by various age groups. It would be nice to have income and other demographic variables.

I obtained the road network in Toronto from [BBBike](https://download.bbbike.org/osm/bbbike/Toronto), and use {[r5r](https://ipeagit.github.io/r5r/index.html)} to calculate driving times from DA centroids to each clinic. 

The file with the road network is not shared on GitHub (it is a large file), so if you wish to replicate the routing calculations you need to obtain a copy and place in folder `data-raw/r5_graph/`. The file must be in `osm.pbf` format, which is what {r5r} uses.

The notebook with the routing is in folder `data-raw\01-OSM-Network-and-Routing`. You can check it for details.

The following data objects are available:

- `data/clinics.rda`: a simple features table with the location of the clinics.  
- `data/da_centroids.rda`: a simple features table with the centroids of the Dissemination Areas (DAs) in Toronto.
- `data/ttm_driva_da.rda`: a simple features table with Toronto's Dissemination Areas (DAs) and population statistics by various age groups.  
- `data/ttm_driva_da.rda`: a data frame with driving times from DA centroid to clinic. The travel time is in minutes.

## Calculate accessibility and availability

Load packages:
```{r load-packages, include = TRUE}
library(accessibility) # Transport Accessibility Measures
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(sf) # Simple Features for R
library(tidyr) # Tidy Messy Data
```

Load data:
```{r load-data, include = TRUE}
load("data/clinics.rda")
load("data/ttm_drive_da.rda")
load("data/toronto_das.rda")
```

Calculate the impedance, using a step-wise function:
```{r calculate-impedance, include = TRUE}
# Calculate impedance; use the formula used by Paez, Higgins, and Vivona (2019). Check equation in page 27.
ttm_drive_da <- ttm_drive_da |>
  mutate(f_tt = case_when(travel_time <= 5 ~ 0.946,
                          travel_time > 5 & travel_time <= 10 ~ 0.801,
                          travel_time > 10 & travel_time <= 15 ~ 0.607,
                          travel_time > 15 & travel_time <= 20 ~ 0.411,
                          travel_time > 20 & travel_time <= 30 ~ 0.135,
                          travel_time > 30 & travel_time <= 45 ~ 0.011,
                          travel_time > 45 ~ 0.000))
```

The accessibility is the weighted sum of the clinics, with the weights given by the impedance:
```{r calculate-accessibility, include = TRUE}
acc_sex <- ttm_drive_da |>
  group_by(DAUID) |>
  summarize(S = sum(f_tt),
            .groups = "drop")
```

Summary of the accessibility:
```{r summary-accessibility, include = TRUE}
summary(acc_sex)
```

Join the accessibility to the DAs:
```{r join-accessibility-to-DAs, include = TRUE}
toronto_das <- toronto_das |>
  left_join(acc_sex,
            by = "DAUID")
```

Plot the accessibility:
```{r plot-accessibility, include = TRUE}
ggplot() +
  geom_sf(data = toronto_das,
          aes(fill = S),
          color = NA) +
  scale_fill_distiller(name = "Accessibility",
                       palette = "Reds", 
                       direction = 1,
                       na.value = "white") +
  labs(subtitle = "Accessibility to sex health clinics by DA") +
  theme_minimal()
```

Prepare a table with the population and clinics to use as a land use table in package {accessibility}:
```{r prepare-table-for-spatial-availability, include = TRUE}
# Prepare land use table
# Population
pop <- toronto_das |>
  st_drop_geometry() |>
  transmute(id = DAUID, 
            population = Pop15plus,
            clinics = 0)

# Clinics
opps <- clinics |>
  st_drop_geometry() |>
  transmute(id = id, 
            population = 0,
            clinics = 1)

# Bind into a single "land use" table
lu <- rbind(pop, opps)
```

Calculate the spatial availability of clinics (detailed table is the portion of clinic allocated to each DA):
```{r calculate-detailed-spatial-availability, include = TRUE}
detailed_V <- spatial_availability(
  ttm_drive_da |>
    transmute(from_id = DAUID, 
              to_id = id,
              travel_time),
  lu |>
    mutate(population = replace_na(population, 0)),
  opportunity = "clinics",
  travel_cost = "travel_time",
  demand = "population",
  decay_function = decay_stepped(c(5, 10, 15, 20, 30, 45),
                                 weights = c(0.801, 0.607, 0.411, 0.135, 0.011, 0)),
  detailed_results = TRUE
)
```

Calculate the total spatial availability by DA (the sum of all portions of clinics allocated to each DA):
```{r calculate-spatial-availability, include = TRUE}
V <- detailed_V |>
  group_by(from_id) |>
  summarize(V = sum(clinics),
            .groups = "drop") |>
  rename(DAUID = from_id)
```

Join spatial availability to DAs:
```{r join-spatial-availability-to-DAs, include = TRUE}
toronto_das <- toronto_das |>
  left_join(V,
            by = "DAUID")
```

Calculate the availability per capita $v_i$, this is the availability divided by the population serviced in the DA:
```{r calculate-availability-per-capita, include = TRUE}
toronto_das <- toronto_das |>
            mutate(v = V/Pop15plus)
```

Compare the accessibility and availability:
```{r compare-accessibility-and-availability, include = TRUE}
S_plot <- ggplot() +
  geom_sf(data = toronto_das,
          aes(fill = S),
          color = NA) +
  scale_fill_distiller(name = "Accessibility",
                       palette = "Reds", 
                       direction = 1,
                       na.value = "white") +
  labs(subtitle = "Accessibility to sex health clinics by DA") +
  theme_minimal()

V_plot <- ggplot() +
  geom_sf(data = toronto_das,
          aes(fill = V),
          color = NA) +
  scale_fill_distiller(name = "Availability",
                       palette = "Reds", 
                       direction = 1,
                       na.value = "white") +
  labs(subtitle = "Availability of sex health clinics by DA") +
  theme_minimal()

S_plot
V_plot
```

Plot the availability per capita:
```{r plot-availability-per-capita, include = TRUE} 
ggplot() +
  geom_sf(data = toronto_das,
          aes(fill = v),
          color = NA) +
  scale_fill_distiller(name = "Availability per capita",
                       palette = "Reds", 
                       direction = 1,
                       na.value = "white") +
  labs(subtitle = "Availability per capita to sex health clinics by DA") +
  theme_minimal()
```

Plot the deviations of the availability per capita relative to the equality line (the number of clinics divided by the total population serviced):
```{r equity-plot, include = TRUE}
# Equal proportion is clinics divided by population
ep <- 15/sum(toronto_das$Pop15plus, na.rm = TRUE)
toronto_das <- toronto_das |>
  mutate(dev_v = v - ep)

# Plot deviations from equal distribution
ggplot() +
  geom_sf(data = toronto_das,
          aes(fill = dev_v),
          color = NA) +
  scale_fill_gradient2(name = "Availability per capita",
                       #palette = "Reds", 
                       #direction = 1,
                       na.value = "white") +
  labs(subtitle = "Availability per capita to sex health clinics by DA") +
  theme_minimal()
```

```{r}
toronto_das |>
  st_drop_geometry() |>
  select(Median_Income,
         S,
         V) |>
  pivot_longer(-Median_Income,
               names_to = "score",
               values_to = "value") |>
  ggplot(data = ) +
  geom_point(aes(x = Median_Income,
                 y = value)) +
  theme_minimal() +
  facet_wrap(~ score, scales = "free_y")
```

```{r}
cor(toronto_das$Median_Income, 
    toronto_das$S, 
    use = "pairwise.complete.obs")

cor(toronto_das$Median_Income, 
    toronto_das$V, 
    use = "pairwise.complete.obs")
```

```{r}
toronto_das |>
  ggplot() +
  geom_point(aes(x = Median_Income,
                 y = dev_v)) +
  theme_minimal()
```

```{r}
cor(toronto_das$Median_Income, 
    toronto_das$dev_v, 
    use = "pairwise.complete.obs")
```



```{r}
mod <- lm(dev_v ~ Median_Income + Low_Income + Average_Age + Prop_Visible_Minority + Prop_Indigenous_Pop + Prop_One_Parent_Families + Prop_Female,
   data = toronto_das |>
     st_drop_geometry() |>
     mutate(Median_Income = Median_Income/10000))

summary(mod)
```

```{r}
mod <- lm(dev_v ~ Median_Income + Low_Income + Average_Age + Prop_Female,
   data = toronto_das |>
     st_drop_geometry() |>
     mutate(dev_v = dev_v/ep,
            Median_Income = Median_Income/10000))

summary(mod)
```