---
title: "Travel time calculations"
output: html_notebook
---

In this notebook we calculate travel times between DA centroids and clinics. The source of the clinic data is...ASK EVA BOONSMA...and the centroids are...ALSO ASK.

I used [BBBike](https://download.bbbike.org/osm/bbbike/) to extract OSM data for Hamilton on April 16, 2024. The name of the file is `Toronto.osm.pbf`. I copied this file to folder `data-raw/r5_graph`.

IMPORTANT: allocate memory before loading {r5r}. The memory needs to be consistent with the system where this is run. In AP's laptop this is set to 20 Gigabytes
```{r}
# Allocate memory for routing
options(java.parameters = "-Xmx20G")
```

Load packages:
```{r}
library(disk.frame) # Larger-than-RAM Disk-Based Data Manipulation Framework
library(dplyr) # A Grammar of Data Manipulation
library(ggforce) # Accelerating 'ggplot2'
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(glue) # Interpreted String Literals
library(here) # not installed on this machine
library(purrr) # Functional Programming Tools
library(r5r) # Rapid Realistic Routing with 'R5'
library(readr) # Read Rectangular Text Data
library(sf) # Simple Features for R
library(skimr) # Compact and Flexible Summaries of Data
library(stringr) # Simple, Consistent Wrappers for Common String Operations
library(units) # Measurement Units for R Vectors
```

Note that {r5r} requires the Java SE Development Kit 21. See:

https://www.oracle.com/java/technologies/downloads/#java21

Set the environmental variable to use Java: https://confluence.atlassian.com/doc/setting-the-java_home-variable-in-windows-8895.html

## Read data

There are two data files with locational information, centroids of DAs and location of clinics. Read the files:
```{r}
clinics <- read_csv("clinics.csv")
das <- read_csv("centroids.txt")
```

Convert to simple features:
```{r}
das <- das |>
  st_as_sf(coords = c("X", 
                      "Y"), crs = 32617) |>
  st_transform(crs = 4326)

clinics <- clinics |>
  # Convert to simple features
  st_as_sf(coords = c("Longitude", 
                      "Latitude"), crs = 4326) |>
  # Make an id for the clinics
  mutate(id = 1:n(),
         id = as.character(id),
         id = case_when(str_length(id) == 1 ~ paste0("C0", id),
                        TRUE ~ as.character(id))) |>
  # Rename the columns
  rename(Clinic_Name = `Clinic Nam`,
         Location = `Clinic Loc`) |>
  # Reorganize the columns in the table
  select(id, everything())
```

## Network routing

Set Up R5 Routing. First define the path to where the graph is located:
```{r set up r5 path, include=FALSE}
r5_path <- file.path("./r5_graph")
```

This is an example of downloading GTFS (general transit feed specification) for Hamilton, needs to be changed for Toronto if desired (NOT RUN):
```{r eval=FALSE}
# download.file(url = "https://transitfeeds.com/p/hamilton-street-railway/31/latest/download", 
#               destfile = file.path(r5_path, 
#                                    "HSR_transit.zip"), 
#               mode = "wb")
```

Build the graph:
```{r build-graph, include = FALSE}
r5_toronto <- setup_r5(data_path = r5_path, 
                       verbose = FALSE)
```

Retrieve network as a simple features object:
```{r eval=FALSE}
toronto_net <- street_network_to_sf(r5_toronto)
```

Save network as sf:
```{r eval=FALSE}
save(toronto_net, file = "hamilton_net.rda", compress = "xz")
```

Prepare Input Data for `r5r` to calculate travel times between DA centroids and clinics. 

The origins are the coordinates of the DAs and the destinations the coordinates of the clinics:
```{r prepare-inputs}
# save origins in format expected by R5R (id, lon, lat)
origins_i_da <- data.frame(id = das$DAUID, 
                           das |>
                             st_transform(crs = 4326) |>
                             st_coordinates()) |>
  rename(lon = X, lat = Y) |>
  dplyr::select(id, lon, lat)

# now destinations sites
destinations_j_clinics <- data.frame(id = clinics$id, 
                                     clinics |>
                                       st_transform(crs = 4326) |>
                                       st_coordinates()) |>
  rename(lon = X, lat = Y) |>
  dplyr::select(id, lon, lat)
```

Calculate OD Matrix for driving (from DAs to clinics):
```{r calculate drive od matrix DAs}

# set up batching according to how many origin rows to process at one time
chunksize = 4000 # larger chunks for walking will give enough origins in each chunk to allow multiprocessing to spin up with R5R
num_chunks = ceiling(nrow(origins_i_da)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(origins_i_da,
                                outdir = "./df/origins_i_da",
                                nchunks = num_chunks,
                                overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){ 
  origins_i_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5_toronto,
                                  origins = origins_i_chunk,
                                  destinations = destinations_j_clinics,
                                  mode = c("CAR"),
                                  departure_datetime = as.POSIXct(strptime("2024-04-16 08:00:00", "%Y-%m-%d %H:%M:%S")),
                                  max_walk_time = 15, # minutes
                                  max_trip_duration = 150)
  
  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "./df/output_ttm_drive",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}

end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```

Time taken:
```{r}
time.taken
```


## Extract travel time matrices

For DAs:
```{r load od matrix for walk DAs, include=FALSE}
# connect to the walking travel time matrix disk frame
ttm_drive.disk.frame <- disk.frame("./df/output_ttm_drive")
#opportunities_j <- as.data.frame(nyc_cb_point) |> select(GEOID10, total_emp) |> rename(toId = GEOID10, o_j = total_emp)
```

Convert disk.frame to data frame:
```{r}
ttm_drive_da <- as.data.frame(ttm_drive.disk.frame) |>
  transmute(DAUID = from_id, id = to_id, travel_time = travel_time_p50)
```

Summary of travel time table:
```{r}
summary(ttm_drive_da)
```

Save data objects:
```{r eval=FALSE}
# DA centroids
da_centroids <- das

da_centroids <- da_centroids |>
  st_transform(crs = 4326)

save(da_centroids,
     file = glue(here::here(), 
                 "/data/da_centroids.rda"), 
     compress = "xz")

# Clinics
save(clinics,
     file = glue(here::here(), 
                 "/data/clinics.rda"), 
     compress = "xz")

# Travel time matrix
save(ttm_drive_da, file = glue(here::here(), "/data/ttm_drive_da.rda"), compress = "xz")
```


